A Data Restore Model\\ for Reproducibility in Computational Statistics

### 2013-05-07 paper-data-restore-model.Rnw
cleanups

\newcommand{\conceptual}[1]{
%\textcolor{RoyalBlue}{#1}
\texttt{#1}
}

% TODO: wir sollten noch die Terminologie klï¿½ren. wie bezeichnen wir 
%
% * das Prinzip, Werte zu referenzieren:  Bisher nannte ich es Reference Model
%
% * so ein Daten-Template, das IDs anstelle von Werten enthï¿½lt: data reference template? ddocks template?
%
%
% Solange das noch nicht so ganz gefunden ist, kï¿½nnen wir einfach diese Tokens verwenden:
%
\newcommand{\refmodel}{
%data reconstruction model
data restore model
}
\newcommand{\datatemplate}{
data template
}

@@

\title{
%On the Reproducibility of Statistical Research
%On the Reproducibility in Computational Statistics
A Data Restore Model\\
for Reproducibility in Computational Statistics
}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

Researchers are more and more requested to publish their scientific data sets for purposes of transparency, re-use, and reproducibility.
Particularly in economics and the social sciences, researchers often use sensitive statistical data that underlies protection policies that inhibit distribution to third party archives.
In addition, a substantial quantity of data sets is combined from one or more external providers, which easily complicates curation-related activities. %---when working with data copies
These circumstances give us reason to pursue a data restore model based on fine-grained referencing that allows to trace data provenance to the original archive in charge of curation.
One goal is to enable data publication in difficult settings.
%BZ: "this research" koennte verwechselt werden mit der research, die reproduzierbar sein soll. wie waers mit "our goal is to enable ..."
Moreover, we show how the gaps between data citation and code integration can be closed as to eliminate all manual efforts of arranging code and data for reproduction attempts. 
On this basis, we develop the requirements for a data restore model and elaborate a generic design in view of an overall data management infrastructure.
We further present an experimental implementation on the basis of a real-world example from economics.
Eventually we close with the vision of a data and code ontology that carries statistical models from paper to a re-usable semantic level.

@@

@@

Repeatability of research results is a fundamental criterion in science. % experiments are conducted in different environments and by different persons as to validate whether the results depend on considered parameters only.
Experiments must lead to the same findings when conducted by other persons or in other environments, under the premise of keeping relevant and considered parameters constant.
%This is very often not the case in computational research 
% sciencemag article on "Reproducible research in computational science"
In computational research, this remains unverified in many cases due to various reasons.
\cite{peng2011reproducible}
%One of them is the plain fact that the data on which the research is built often remains unavailable to others which gives rise for a more organized data management practice. \cite{Wood1298248}
One of them is the plain fact that the supporting data often is unavailable to others which gives rise for a more organized and open data management practice. \cite{Wood1298248}
While a scientific publication is commonly understood as synonymous with a peer-reviewed article alone, program code and data take the role of nice-to-have additional material only that rarely find their way to the public.
%There are numerous examples of research publications that turned out inconsistent eventually, either because of mistakes,  
%There are numerous examples of inconsistent research  that illustrate how

@@

Both, data and code resources make a valuable contribution if findings are to be reproducible and re-usable in computational research.
And yet, there is more to be considered.
%A comprehensive review of the JMCB\footnote{Journal of Money, Credit, and Banking} data+code archive reveals that results cannot be reproduced even when these resources are given.  
A comprehensive review of the \textit{JMCB\footnote{Journal of Money, Credit, and Banking} data+code archive} reveals that provenance information for the data used is often incomplete and that submitted code does not reproduce the results either.
It is not uncommon that even the original authors have difficulties understanding the details of their own work only few months after publishing already.
Apparently, these issues are faced in many other areas of computational research too
% reproducibility is a problem in many domains
% bio informatics (case studies 100 articles..) mit leisch, sweave autor
\cite{hothorn2011case}
% empirical software engineering
\cite{gonzalez2012reproducibility}
% signal processing
\cite{vandewalle2009reproducible}
% ecology
\cite{cassey2006reproducibility}
% "replicating empirical results is an underemphasized, yet essential element of scientific progress."
% Report on the American Economic Review Data Availability Compliance Project
% http://www.aeaweb.org/aer/2011_Data_Compliance_Report.pdf
which can be understood as a problem of insufficient documentation.

@@

However, empirical research is often based on sensitive data that cannot be shared easily.
%Such data sets can be transformed into anonymized public-use files in general which needs verification in terms of cell sizes and de-anonymization and requires additional efforts, often on the researchers' side.
A general solution to this problem is to transform such data sets into anonymized public-use files that can be distributed without infringements of data protection.
But doing so requires careful verification in terms of cell sizes and de-anonymization and requires additional efforts, often on the researchers' side.
Moreover, researchers hesitate to publish their own data if incentives are not clear, especially when they had a lot of work producing it.
In addition, data is often obtained from several sources and the mere citation of them does not give any details on the file formats used or way of composition as used for input for the computer program.
%BZ: hier kommst du wieder zurueck zu den daten (wenn auch mit anderem schwerpunkt), nachdem das thema oben eigentlich abgehakt wirkte und du zum thema program code und danach zu provenance gegangen bist. sollte dieser abschnitt ueber sensitive data nicht zu dem data abschnitt weiter oben? gehoert ja irgendwie schon zusammen, oder?

%These circumstances motivate our efforts to elaborate on a holistic solution for the statistical domain that enables the exact reconstruction of individual research data sets as were used by the researchers.
In this paper we aim to overcome these challenges for the domain of statistical data and elaborate on a holistic solution that enables the exact reconstruction of individual research data sets as were used by the researchers.
We pursue the idea of a \emph{data restore model} that is based on fine-grained data referencing as described in earlier works \cite{BahlsTochtermann_2012}
where further background and motivation is explained more thoroughly.
As this referencing technique can be viewed as means for precise citation, the linkup to provenance and further documentation on the particular data version is given which again contributes to clarity.
The idea is to equip researchers with means that make full transparency possible even in difficult settings of protected and distributed data sources, so that reconstruction remains a question of access permission only.
%We aim to design this referencing technique in a way that allows for seamless integration with preferred research tools and source code, so that no further manual work is required for setting up data and code resources during replication attempts.
%The restore model is designed to seamlessly integrate with preferred research tools and source code, so that no further manual work is required for setting up data and code resources during replication attempts.
The restore model is designed to bridge the gaps between data citation and program code and to seamlessly integrate with preferred research tools, so that no further manual work is required in setting up data and code resources for replication attempts.
Eventually, the goal is to provide researchers with the ability to close all missing links from data sources and processing to figure plotting and report generation as to create trust in their findings which from another perspective could be an incentive to do so.
On the basis of this data restore model, we elaborate the requirements it carries on the entire information infrastructure between researchers, archives, and libraries.

%In this paper we aim to overcome these challenges for the domain of statistical data and elaborate on a holistic solution for an infrastructure that enables reconstruction of individual research data sets. 
%This infrastructure is basically realized by an underlying data reconstruction model, which builds the bridge between the researchers' preferred software tool and the research data. 
%The approach follows the idea of a data reconstruction model by which we develop the requirements it carries on an infrastructure.
%We present an implementation of the model that tries to seamlessly integrate with any software tool preferred by a researcher.
%By utilizing the model generic ids are assigned to all data values regardless of their original format and their physical location.
%This allows for reconstructing and referencing the data set out of the researchers programming tool.

The remainder of the paper is structured as follows. 
In section 2 we present a case study with real world examples, where literate programming with research data is applied. 
Based on these examples we propose a generic data reconstruction model in section 3. 
We implement and validate our model in section 4. 
In section 5 we discuss related work relevant to our approach. 
Finally, we conclude and present future work in section 6.
%BZ: veraltet. muss man am schluss aktualisieren, wenn das paper steht

\section{Related Work}

@@

In this section, we first develop the requirements for a data restore model.
%Subsequently, we elaborate a generic design and outline the big picture of the overall data mangement infrastructure and workflows.
Subsequently, we elaborate a generic design and discuss its implication on an overall data mangement infrastructure and the workflows involved.

@@

Whichever data management solution is applied eventually, it should fit to the researchers' workflows and practices in order to be effective and accepted hence.
\cite{Feijen_2011} 
In the ideal case, researchers can use their preferred tools with no further hassle preparing anything at the time of data submission.
%  (assisting tools to reduce efforts)
%
%On the side of reuse, there should not be any unnecessary efforts in re-using these resources either, meaning that no puzzling and manual tinkering in organizing data and code to make it run.
On the side of reuse, there should not be any efforts needed in composing and organizing these resources either, meaning that no puzzling and manual tinkering in organizing data and code to make it run is required \textit{(F partly)}.
In the following, we want to develop the ideas of a referencing technique and take into account the recommendations \textit{(A-J)} fromulated by \cite{RePEc_ejw_journl_v_4_y_2007_i_3_p_326_337} which we refer to using angle brackets.
Location and structure of data files on the local file system should be inherently specified, so that it integrates perfectly with the code.
\footnote{In this regard, software and system specifications are as important. This kind of documentation can be organized independent of the issues addressed in this paper, we therefore do not further address it here.}

@@

% zu weich, was heißt das?  
%  \item overall, it has to meet the constraints and requirements from all parties involved in the overall data management infrastructure, such as archives, researchers and libraries. % was vergessen? und not openly available data = closed data?

@@

%As this approach is based on URIs, namespaces can be used to identify archives, and in combination with locally organized IDs every single value of any data set could be referenced from any context (\textit{Tracability}).
%The requirement item \textit{Curation} is addressed as well, because data can be referenced with precision and maintenance of every data can done in one place. 
%Since the use of RDF allows us to add an additional link between ID and value, we can separate the two and disclose all information about the data in very detail in machine-understandable form, and yet protect the values (\textit{Protection}).
%Furthermore, when every value can be referred to from anywhere, a simple technique can be applied to enable format-independent reconstruction of data files.
%We therefore introduce the concept of a \textit{data template} which is based on the idea to replace every value within a data file with its associated ID, so that the original file can be restored again simply by replacing them back with its associated values which solves the requirements of \textit{Equality}.
%The data template contains a header section in which we can specify encoding and tokens that are used to identify IDs within the template.
%It enables the format-independent solution (\textit{Formats}) as this flexibility allows us to avoid syntactical clashes with the particular data format used.
%In addition, we also declare Web Service addresses for every namespace used (\textit{API}) under the pragmatic assumption that every archive provides one such Web Service for data enquiries only.
%At this point, we do not want to elaborate on the Web Service interface, however, we have used SPARQL endpoints in the proof of concept implementation presented in Section \ref{sec:impl}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=.48\textwidth]{img/fig-model-processing.png}
\caption{
Using the \refmodel
}
\label{fig:modelprocessing}
\end{figure}

@@

\begin{figure*}[htb]
\centering
\includegraphics[width=.95\textwidth]{img/datatemplate.png}
\caption{
Excerpt from one of the data templates
}
\label{fig:datatemplate}
\end{figure*}

@@

%
%
%\begin{quote}

\url{https://github.com/dbahls/ddocks-utils.git}
%\end{quote}

%For a more detailed view on the implementation, please take a look at the source code.\footnote{\url{https://github.com/dbahls/ddocks} \url{}}

@@

<<results=hide>>= 
ddocks_restore(
  file("./data-templates/rk.raw.ddocks"), 
  targetfile="rk.raw")
@
and reproduce the data files on the local file system.
\textit{ddocks} takes these files as a cache and skips when the restore command is run again, so that the data does not have to be reconstructed every time the code is executed.
%Researchers could embed these calls within their source code to load the data by template at the beginning of their analyses and leave it there without being bothered.
Researchers could embed these calls within their source code to load the data by template at the beginning of their analyses without being bothered, and no further modifications are needed at the time of publishing.
%This allows us to comfortably use the code within the source code of a research project from the begin
%
%
%\begin{quote}

\url{https://github.com/dbahls/ddocks.git}
%\end{quote}

@@

We also inserted the two restore calls above in the source code of our example paper (\cite{koenkerzeileis09}) right before the load-data calls respectively, and running Sweave on it produced a report identical to the published article.\footnote{using R x64 2.15.3 on Windows}
%
%
%\begin{quote}

\url{https://github.com/dbahls/paper-restore-model.git}
%\end{quote}

\section{Summary, Vision and Outlook}

>>>>

\newcommand{\datatemplate}{
data template
}

@@

\title{
A Data Restore Model\\
for Reproducibility in Computational Statistics
}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

Researchers are more and more requested to publish their scientific data sets for purposes of transparency, re-use, and reproducibility.
Particularly in economics and the social sciences, researchers often use sensitive statistical data that underlies protection policies that inhibit distribution to third party archives.
In addition, a substantial quantity of data sets is combined from one or more external providers, which easily complicates curation-related activities. %---when working with data copies
These circumstances give us reason to pursue a data restore model based on fine-grained referencing that allows to trace data provenance to the original archive in charge of curation.
One goal is to enable data publication in difficult settings.
Moreover, we show how the gaps between data citation and code integration can be closed as to eliminate all manual efforts of arranging code and data for reproduction attempts. 
On this basis, we develop the requirements for a data restore model and elaborate a generic design in view of an overall data management infrastructure.
We further present an experimental implementation on the basis of a real-world example from economics.
Eventually we close with the vision of a data and code ontology that carries statistical models from paper to a re-usable semantic level.

@@

\newpage

@@

Repeatability of research results is a fundamental criterion in science. % experiments are conducted in different environments and by different persons as to validate whether the results depend on considered parameters only.
Experiments must lead to the same findings when conducted by other persons or in other environments, under the premise of keeping relevant and considered parameters constant.
% sciencemag article on "Reproducible research in computational science"
In computational research, this remains unverified in many cases due to various reasons.
\cite{peng2011reproducible}
One of them is the plain fact that the supporting data often is unavailable to others which gives rise for a more organized and open data management practice. \cite{Wood1298248}
While a scientific publication is commonly understood as synonymous with a peer-reviewed article alone, program code and data take the role of nice-to-have additional material only that rarely find their way to the public.
%There are numerous examples of research publications that turned out inconsistent eventually, either because of mistakes,  
%There are numerous examples of inconsistent research  that illustrate how

@@

Both, data and code resources make a valuable contribution if findings are to be reproducible and re-usable in computational research.
And yet, there is more to be considered.
A comprehensive review of the \textit{JMCB\footnote{Journal of Money, Credit, and Banking} data+code archive} reveals that provenance information for the data used is often incomplete and that submitted code does not reproduce the results either.
It is not uncommon that even the original authors have difficulties understanding the details of their own work only few months after publishing already.
Apparently, these issues are faced in many other areas of computational research too
% reproducibility is a problem in many domains
% bio informatics (case studies 100 articles..) mit leisch, sweave autor
\cite{hothorn2011case}
% empirical software engineering
\cite{gonzalez2012reproducibility}
% signal processing
\cite{vandewalle2009reproducible}
% ecology
\cite{cassey2006reproducibility}
% "replicating empirical results is an underemphasized, yet essential element of scientific progress."
% Report on the American Economic Review Data Availability Compliance Project
% http://www.aeaweb.org/aer/2011_Data_Compliance_Report.pdf
which can be understood as a problem of insufficient documentation.

@@

However, empirical research is often based on sensitive data that cannot be shared easily.
A general solution to this problem is to transform such data sets into anonymized public-use files that can be distributed without infringements of data protection.
But doing so requires careful verification in terms of cell sizes and de-anonymization and requires additional efforts, often on the researchers' side.
Moreover, researchers hesitate to publish their own data if incentives are not clear, especially when they had a lot of work producing it.
In addition, data is often obtained from several sources and the mere citation of them does not give any details on the file formats used or way of composition as used for input for the computer program.
%BZ: hier kommst du wieder zurueck zu den daten (wenn auch mit anderem schwerpunkt), nachdem das thema oben eigentlich abgehakt wirkte und du zum thema program code und danach zu provenance gegangen bist. sollte dieser abschnitt ueber sensitive data nicht zu dem data abschnitt weiter oben? gehoert ja irgendwie schon zusammen, oder?

%These circumstances motivate our efforts to elaborate on a holistic solution for the statistical domain that enables the exact reconstruction of individual research data sets as were used by the researchers.
In this paper we aim to overcome these challenges for the domain of statistical data and elaborate on a holistic solution that enables the exact reconstruction of individual research data sets as were used by the researchers.
We pursue the idea of a \emph{data restore model} that is based on fine-grained data referencing as described in earlier works \cite{BahlsTochtermann_2012}
where further background and motivation is explained more thoroughly.
As this referencing technique can be viewed as means for precise citation, the linkup to provenance and further documentation on the particular data version is given which again contributes to clarity.
The idea is to equip researchers with means that make full transparency possible even in difficult settings of protected and distributed data sources, so that reconstruction remains a question of access permission only.
The restore model is designed to bridge the gaps between data citation and program code and to seamlessly integrate with preferred research tools, so that no further manual work is required in setting up data and code resources for replication attempts.
Eventually, the goal is to provide researchers with the ability to close all missing links from data sources and processing to figure plotting and report generation as to create trust in their findings which from another perspective could be an incentive to do so.
On the basis of this data restore model, we elaborate the requirements it carries on the entire information infrastructure between researchers, archives, and libraries.

The remainder of the paper is structured as follows. 
In Section \ref{sec:relatedwork} we discuss related work relevant to our approach. 
%In section 2 we present a case study with real world examples, where literate programming with research data is applied. 
%Based on these examples we propose a generic data reconstruction model in section 3. 
The requirements for a data restore model are elaborated in Section \ref{sec:datareferencing} by which we develop a generic solution that.
An experimental implementation is presented in Section \ref{sec:impl} where we explore its application in context of a case study from economics.
%Finally, we conclude and present future work in Section \ref{sec:outlook}.
Finally, we conclude in Section \ref{sec:outlook} with summary and outlook on future work and present a vision that illustrates the potential of this approach .

\section{Related Work}\label{sec:relatedwork}

@@

In this section, we first develop the requirements for a data restore model.
Subsequently, we elaborate a generic design and discuss its implication on an overall data mangement infrastructure and the workflows involved.

@@

In the following, we want to develop the ideas of a referencing technique and take into account the recommendations \textit{(A-J)} fromulated by \cite{RePEc_ejw_journl_v_4_y_2007_i_3_p_326_337} which we refer to using angle brackets.
Whichever data management solution is applied eventually, it should fit to the researchers' workflows and practices in order to be effective and accepted hence.
\cite{Feijen_2011} 
In the ideal case, researchers can use their preferred tools with no further hassle preparing anything at the time of data submission.
%  (assisting tools to reduce efforts)
%
%On the side of reuse, there should not be any unnecessary efforts in re-using these resources either, meaning that no puzzling and manual tinkering in organizing data and code to make it run.
On the side of reuse, there should not be any efforts needed in composing and organizing these resources either, meaning that no puzzling and manual tinkering in organizing data and code to make it run is required \textit{(F partly)}.
Location and structure of data files on the local file system should be inherently specified, so that it integrates perfectly with the code.
\footnote{In this regard, software and system specifications are as important. This kind of documentation can be organized independent of the issues addressed in this paper, we therefore do not further address it here.}

@@

%  \item overall, it has to meet the constraints and requirements from all parties involved in the overall data management infrastructure, such as archives, researchers and libraries. % was vergessen? und not openly available data = closed data?

@@

@@

\begin{figure}[htb]
\centering
\includegraphics[width=.48\textwidth]{img/fig-model-processing.png}
\caption{
Using the data restore model
}
\label{fig:modelprocessing}
\end{figure}

@@

\begin{figure*}[htb]
\centering
\includegraphics[width=.95\textwidth]{img/datatemplate.png}
\caption{
%Excerpt from one of the data templates
%Data template excerpt from one of the data templates
Data template for one of the two data sets (\textit{rk.raw})
}
\label{fig:datatemplate}
\end{figure*}

@@

%
% link to source code
%
\url{https://github.com/dbahls/ddocks-utils.git}

@@

<<results=hide>>= 
ddocks_restore(
  file("./data-templates/rk.raw.ddocks"), 
  targetfile="rk.raw")
@
and reproduce the data files on the local file system.
\textit{ddocks} takes these files as a cache and skips when the restore command is run again, so that the data does not have to be reconstructed every time the code is executed.
%Researchers could embed these calls within their source code to load the data by template at the beginning of their analyses and leave it there without being bothered.
Researchers could embed these calls within their source code to load the data by template at the beginning of their analyses without being bothered, and no further modifications are needed at the time of publishing.

%
% link to source code
%
\url{https://github.com/dbahls/ddocks.git}

@@

We also inserted the two restore calls above in the source code of our example paper (\cite{koenkerzeileis09}) right before the load-data calls respectively, and running Sweave on it produced a report identical to the published article.\footnote{using R x64 2.15.3 on Windows}

%
% link to source code
%
\url{https://github.com/dbahls/paper-restore-model.git}

\section{Summary, Vision and Outlook}\label{sec:outlook}
